---
title: "Team 05"
subtitle: "Proposal"
format: html
editor: visual
---

```{r}
#| label: load-pkgs
#| message: false

library(tidyverse)
```

# Data 1

## Introduction and data

[2021 Students In Temporary Housing - Catalog (data.gov)](https://catalog.data.gov/dataset/2021-students-in-temporary-housing)

This data was collected from the Data.Gov Data Catalog. This data set was created on December 11, 2021, and the data was first published on December 6, 2021. The data is sourced from the City of New York. There is not much information concerning the specific method of collecting data; however, the data was collected for the New York City Department of Education, who likely used surveys to find out the relevant information about students who attended New York City district schools from the beginning of the school year to July 2 over the course of the 2020-2021 school year.

Each observation is based on a DBN (District Borough Number) and a school's name. For each observation, there are different variables to identify the number of total students, the percentage of students in temporary housing, the number of students residing in shelter, the number of students residing in a DHS (Department of Homeless Services) shelter, the number of students residing in a non-DHS shelter, and the number of students living in "doubled up" housing, which means that they are living in households with multiple families due to the financial inability to live in their own home.

## Research question

Does whether or not a student attended a public school during the 2020-2021 school year impact the likelihood of that student to be how likely that student was to experience housing instability during that year?

Did students who attend public schools experience different types of housing instability than those who attend private schools during the 2020-21 school year?

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

Our hypothesis is that students who attended a public school during the 2020-21 school year had a higher likelihood of experiencing housing instability than their privately-educated peers.

We also hypothesize that students who are educated at private schools were more likely to experience doubling up housing insecurity rather than living in shelters. On the other hand, I hypothesize that students educated at public schools experienced living in shelters as a a form of housing instability at a higher rate than those who were educated in private schools.

There are both quantitative and categorical variables in my research question:

The categorical variable for our first research question is type of education (private or public), and the quantitative variable for our first research question is percent of students in temporary housing.

The categorical variable for our second research question is type of education (private or public), and the quantitative variables for our second research question are \# of students residing in shelter and \# doubled up.

## Literature

The New York Times article, published on October 26th, 2022, explores the challenges faced by homeless students in New York City's public schools. According to the article, there are over 111,000 homeless students enrolled in the city's public schools, a number that has swelled from roughly 78,000 a decade ago. Closson writes about the circumstances that the students face, including inadequate housing, lack of access to food and medical care, and high levels of stress and trauma. It is a huge challenge to provide support and resources to these students, given limited funding. Despite this, the article also highlights positive efforts being made like "Bridging the Gap," a program that aims to provide academic and emotional support to homeless students. Overall, it paints a complex picture of difficult challenges faced by homeless students in NYC; while there are significant obstacles, there are also efforts underway to make real, tangible change.

The article expands on our leading question of the correlation between public schools and housing insecurity. It details the homeless crisis up close and also provides statistics such as the graduation rate of highschool students living in shelters (60%).

Closson, Troy. "More than 104,000 New York City Students Were Homeless Last Year." The New York Times, The New York Times, 26 Oct. 2022, https://www.nytimes.com/2022/10/26/nyregion/nyc-homeless-students.html?smid=tw-nytimes&smtyp=cur.

## Glimpse of data

```{r}
#| label: load-data-1

students_temp_housing <- read_csv("2021_Students_In_Temporary_Housing.csv", 
                                  show_col_types = FALSE)
glimpse(students_temp_housing)
```

# Data 2

## Introduction and data

[DPD Arrests (UCR NIBRS Reporting) \| Durham, NC Open Data Portal (arcgis.com)](https://live-durhamnc.opendata.arcgis.com/documents/dpd-arrests-ucr-nibrs-reporting/about)

This data is from the Open Data website for the City and County of Durham. This data set has been updated since its original publication, and it was collected by the Durham Police Department. The data set was first published on June 24, 2022 and was more recently updated on January 27, 2023. This data set is updated annually and begins in October 2018. The department collected this data from the FBI's UCR Program Data Collections for the National Incident-Based Reporting System (NIBRS). The data does not go back before October 2018, as the agency used the Summary Reporting System before that date (the data is not comparable across reporting systems).

Each observation is a particular arrest of an adult made starting from October 2018. For each observation, variables include personal descriptors (race, age, ethnicity, sex, etc.), the reason for their arrest, when they were arrested, and the location of their arrest.

## Research question

Is there a correlation between the race and age of the arrested individual and the type of the arrest that is utilized?

Research in the past has been conducted on the impact race has on arrest rates and how there are significant and systemic biases against Black Americans in the criminal justice system. However, this project looks to understand whether there is a correlation between race and the type of arrest. Different types of arrests include warrant-based, on-view, and summoned/cited. Our goal is to determine whether a certain race is more likely to be arrested when a warrant is present or whether a certain race is more likely to be arrested "on-site" (or when a police officer suspects that they had engagement in a criminal act - probable cause for an arrest on-site). We also hope to integrate sex into this question, to examine the how intersectionality may exacerbate biases.

We hypothesize that there would be a correlation between Black men arrested and the fuse of on-site arrests. We also predict that there may be a correlation between White women and the use of warrants to arrest.

For this question, most of the variables will be categorical. We are specifically focusing on three variables in this case, race, sex, and type of arrest, which all serve as categorical variables.

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-2

library(readxl)
durham_arrests <- read_excel("dpd_arrests_(ucr_nibrs_reporting).xlsx")
glimpse(durham_arrests)
```

# Data 3

## Introduction and data

[CORGIS Datasets Project (vt.edu)](https://think.cs.vt.edu/corgis/csv/airlines/)

This data is from the CORGIS Data Set Project. This is the second version of this data set, and it was created by Austin Cory Bart. The data set was created on March 27, 2015. There is little information on how the data was collected; however, it is likely that the statistician took a retrospective approach, looking at past monthly records for different airports across the United States starting in 2003.

Each observation is based on an airport code (shorthand for the airport's full name) and a specific time (year/month). For each observation, there are different variables to identify how many flights were delayed (and for what reason), cancelled, diverted, or on time.

## Research question

At what times of year are there the most flight delays and cancelations, and what is the most common type of reasoning for delays and cancellations at these times?

Flight delays and cancellations often give passengers grief and stress at all different times of the year. However, our question explores whether there is a specific time of year when flights are particularly prone to delays and cancellations. In turn, this would help those looking to fly to know how to best plan for trips that might be impacted by flight changes. Working to understand why such issues occur would also be crucial, both in determining whether certain cancellations or delays can be avoided and in providing some perspective to passengers who have struggled with flight plans in the past.

Based on the past literature on the times of year flights are subject to the most cancellations and delays, we hypothesize that the time most prone to flight cancellations and delays is during the summer months (June through August).

We are utilizing mostly quantitative variables in this research, which the exception of the month data, airline information, and airport information (categorical). Quantitative variables used will include the number of flights delayed and cancelled for a variety of reasons at different airports across the country.

## Literature

Published on July 1st, 2022, the NYT article "Understanding the Summer Air Travel Mess" discusses the challenges faced by travelers during the summer months due to mass delays and cancellations. As a result of being the busiest time of the year, factors such as weather, staffing shortages, and equipment issues have a much more significant impact on schedule disruption. Just last year, low-cost carriers like JetBlue and Spirit canceled 10.3 and 9 percent of flights in April. The article also provides advice on how to minimize the impact of flight delays: book direct flights and choose early morning departures. It's also smart to avoid flying during weekends. Most importantly, though, "never book the last flight of the day." Additionally, Murphy writes about the implications these cancellations have on airlines, damaging them both financially and reputation wise. New technologies may be implemented, but, at "the heart of the issue," airlines respond to supply and demand, increasing flights when people travel more.

The NYT article offers a deeper glimpse into why certain times of the year result in more delays and cancellations. It also provides insight into factors such as departure time and flight (direct) routes.

Murphy, Heather. "Understanding the Summer Air Travel Mess." The New York Times, The New York Times, 1 July 2022, https://www.nytimes.com/2022/07/01/travel/summer-travel-flight-delays-cancellations.html.

## Glimpse of data

```{r}
#| label: load-data-3

airlines <- read_csv("airlines.csv", show_col_types = FALSE)
glimpse(airlines)
```
