---
title: "Team 05"
subtitle: "Proposal"
format: html
editor: visual
---

```{r}
#| label: load-pkgs
#| message: false

library(tidyverse)
```

# Data 1

## Introduction and data

[2021 Students In Temporary Housing - Catalog (data.gov)](https://catalog.data.gov/dataset/2021-students-in-temporary-housing)

This data was collected from the Data.Gov Data Catalog. This data set was created on December 11, 2021, and the data was first published on December 6, 2021. The data is sourced from the City of New York. There is not much information concerning the specific method of collecting data; however, the data was collected for the New York City Department of Education, who likely used surveys to find out the relevant information about students who attended New York City district schools from the beginning of the school year to July 2 over the course of the 2020-2021 school year.

Each observation is based on a DBN (District Borough Number) and a school's name. For each observation, there are different variables to identify the number of total students, the percentage of students in temporary housing, the number of students residing in shelter, the number of students residing in a DHS (Department of Homeless Services) shelter, the number of students residing in a non-DHS shelter, and the number of students living in "doubled up" housing, which means that they are living in households with multiple families due to the financial inability to live in their own home.

## Research question

Does whether or not a student attended a public school during the 2020-2021 school year impact the likelihood of that student to be how likely that student was to experience housing instability during that year?

Did students who attend public schools experience different types of housing instability than those who attend private schools during the 2020-21 school year?

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

My hypothesis is that students who attended a public school during the 2020-21 school year had a higher likelihood of experiencing housing instability than their privately-educated peers.

I also hypothesize that students who are educated at private schools were more likely to experience doubling up housing insecurity rather than living in shelters. On the other hand, I hypothesize that students educated at public schools experienced living in shelters as a a form of housing instability at a higher rate than those who were educated in private schools.

There are both quantitative and categorical variables in my research question:

The categorical variable for our first research question is type of education (private or public), and the quantitative variable for our first research question is percent of students in temporary housing.

The categorical variable for our second research question is type of education (private or public), and the quantitative variables for our second research question are \# of students residing in shelter and \# doubled up.

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-1

students_temp_housing <- read_csv("2021_Students_In_Temporary_Housing.csv")
glimpse(students_temp_housing)
```

# Data 2

## Introduction and data

[DPD Arrests (UCR NIBRS Reporting) \| Durham, NC Open Data Portal (arcgis.com)](https://live-durhamnc.opendata.arcgis.com/documents/dpd-arrests-ucr-nibrs-reporting/about)

-   Identify the source of the data.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

-   Write a brief description of the observations.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

-   Identify the types of variables in your research question. Categorical? Quantitative?

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-2

glimpse(durham_arrests)
```

# Data 3

## Introduction and data

[CORGIS Datasets Project (vt.edu)](https://think.cs.vt.edu/corgis/csv/airlines/)

This data is from the CORGIS Data Set Project. This is the second version of this data set, and it was created by Austin Cory Bart. The data set was created on March 27, 2015. There is little information on how the data was collected; however, it is likely that the statistician took a retrospective approach, looking at past monthly records for different airports across the United States starting in 2003.

Each observation is based on an airport code (shorthand for the airport's full name) and a specific time (year/month). For each observation, there are different variables to identify how many flights were delayed (and for what reason), cancelled, diverted, or on time.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

-   Identify the types of variables in your research question. Categorical? Quantitative?

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-3

glimpse(airlines)
```
